---
output: github_document
bibliography: "mtdesign.bib"
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
# /usr/local/lib/R/site-library/BH/include/boost

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# mtdesign

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/mtdesign)](https://CRAN.R-project.org/package=mtdesign)
<!-- badges: end -->

## Introduction
The package `mtdesign` provides implementations of both Simon [-@SIMON] and Mander & Thompson [-@MANDER]. Other implementations of Simon's methods are available - for example, the `ph2simon` function in the `clinfun` package [@CLINFUN], but these do not provide easy access to non-optimal solutions in the way that `mtdesign` does.  I am not aware of any other R-based implementations of Mander & Thompson's extension to Simon.

## Installation
Once available on CRAN, you can install `mtdesign` in the usual way:

`install.packages("mtdesign")`

You can install the development version of `mtdesign` from [GitHub](https://github.com/openpharma/mtdesign) with:

`devtools::install_github("openpharma/mtdesign")`

## Set up vignette environment
```{r}
# By policy, on CRAN, use only two cores, no matter how many are available.
if (requireNamespace("parallel", quietly = TRUE)) {
  maxCores <- parallel::detectCores()
  maxCores <- ifelse(identical(Sys.getenv("NOT_CRAN"), "true"), maxCores, min(maxCores, 2))
} else {
  maxCores <- 1
}
```

## Example

Suppose that treatments with a response rate of less than 5% are of no interest but those with a response rate of at least 25% are worthy of further development.  A Simon's 2-stage design to seek an efficacy signal with a significance level of 5% and a power of 80% is required. 

```{r example}
library(mtdesign)
library(knitr)
library(dplyr)

simonDesign <- obtainDesign(p0 = 0.05, p1 = 0.25, alpha = 0.05, beta = 0.2, mander = FALSE, parallel = FALSE)

simonDesign %>%
  select(-Alpha, -Beta, -p0, -p1, -PETAlt, -AveSizeAlt) %>%
  kable(digits = c(0, 0, 0, 0, 3, 3, 2, 1, NA))
```

The table shows that the optimal design for these requirements is 0/9 2/17.  The expected sample size is 12.0 and the probability of early termination is 63%. The significance level actually achieved is 4.7% and the power level achieved is 100% - 18.8% = 81.2%.

The power curves for both designs are easily plotted.

```{r}
powerPlot(simonDesign)
```

Obtaining the equivalent Mander & Thompson designs requires only a small change to the calls.

```{r}
manderDesign <- obtainDesign(
                  p0 = 0.05, 
                  p1 = 0.25, 
                  alpha = 0.05, 
                  beta = 0.2, 
                  cores = maxCores
                )

manderDesign %>%
  select(-Alpha, -Beta, -p0, -p1) %>%
  kable(digits = c(0, 0, 0, 0, 3, 3, 2, 2, 2, 1, NA))

powerPlot(manderDesign)
```

### Constrained designs

Suppose a trial, for whatever reason, is restricted to using 8 participants in each stage.  As shown above, the optimal Simon's two stage design is 0/9 2/17.  That's close to n~1~ = 8, n = 16.  Is there a (slightly) sub-optimal design that has n~1~ = 8, n = 16?

```{r}
x <- createGrid(p0 = 0.05, p1 = 0.25, alpha = 0.05, beta = 0.2, mander = FALSE)

y <- x %>% filter(nStage1 == 8, nTotal == 16)
z <- y %>% obtainDesign(cores = maxCores)
if (nrow(z) == 0) {
  print("No acceptable designs were found.")
} else {
  select(-Alpha, -Beta, -p0, -p1, -PETAlt, -AveSizeAlt) %>%
    z() %>%
    select(-Alpha, -Beta, -p0, -p1, -PETAlt, -AveSizeAlt) %>%
    kable(digits = c(0, 0, 0, 0, 3, 3, 2, 1, NA))
}
```

No, there isn't.  How close can we get?

```{r}
z1 <- y %>% augmentGrid()

bestSize <- z1 %>%
  filter(Type1 < Alpha) %>%
  slice_min(Type2)
bestSize %>%
  select(-Alpha, -Beta, -p0, -p1, -PETAlt, -AveSizeAlt) %>%
  kable(
    caption = "Best sub-optimal design with required significance level",
    digits = c(0, 0, 0, 0, 3, 3, 2, 1, NA)
  )

bestPower <- z1 %>%
  filter(Type2 < Beta) %>%
  slice_min(Type1)

bestPower %>%
  select(-Alpha, -Beta, -p0, -p1, -PETAlt, -AveSizeAlt) %>%
  kable(
    caption = "Best sub-optimal design with required power",
    digits = c(0, 0, 0, 0, 3, 3, 2, 1, NA)
  )
```

So the choice lies between a design which achieves the required significance level but has a power of only 77.1% or one which has the required power but which has a significance level of 15.1%.  Both designs accept the null hypothesis when no responders are seen in the first group of eight participants.  They differ in the critical value at the end of stage 2: 1 to maintain the power, 2 to maintain the significance level.

The power curve for each of these designs can be compared with that for the globally optimal design.

```{r}
plotData1 <- simonDesign %>%
  filter(Criterion == "optimal") %>%
  bind_rows(list(bestSize, bestPower))
powerPlot(plotData1)
```

## Package structure
The `mtdesign` package consists of three main functions:

* `createGrid` creates the grid (of nStage1, rFutility, nTotal and rTotal for  Simon's design or nStage1, rFutility, rSuccess, nTotal and rTotal for a Mander & Thompson design) over which the brute force search for the required design(s) is conducted
* `augmentGrid`takes a grid created by `createGrid` and adds columns for probability of early termination, Type 1 error, Type 2 error and expected sample size to it.
* `obtainDesign` takes an augmented grid and identifies the optimal and minimax designs

## Error and warning messages and logging
The `mtdesign` package supports logging via the `logger` package [@LOGGER].  Most functions simply report Entry and Exit at the `DEBUG` level.  

The `augmentGrid` function reports steps of the parallelisation process at the `TRACE` level.

## Parallelisation
There is no known closed form solution to obtaining solutions to either Simon's original equations nor Mander & Thompson's extensions.  The `mtdesign` package uses a brute force approach to evaluate the operating characteristics of all reasonable potential designs.  The grids can be quickly become large, particularly for Mander & Thompson designs.  For example, `createGrid(0.2, 0.4, alpha=0.1, beta=0.1)` creates a grid of almost 11 million candidate designs.  `mtdesign` uses paralellisation to attempt to speed up the evaluation of candidate designs.  

The `augmentGrid` function allows users some control over the parallelisation process:

*  The `parallel` parameter defaults to `TRUE` and defines whether or not paralellisation is to be used.
*  The `cores` parameter specifies how many cores are to be used.  The default value, `NA` tells `mtdesign` to use all available (as defined by `parallel::detectCores()`), cores.
*  The `minChunkSize` determines the smallest grid of candidate designs that will trigger paralellisation.  The default value is `100000`.

The `parallel` package is required for parallelisation.  If parallelisation is both needed (ie the grid size exceeds `minChunkSize`) and requested but the `parallel` package has not been installed, an error message is thrown and augmentation of the grid stops.  If paralellisation is not requested and the grid contains one million or more rows, a warning is produced.

## Troubleshooting
If, when installing or using the `mtdesign` package, you get an error regarding a syntax error in an`.hpp` file, similar to the following

```R
.../BH/include/boost/math/tools/fraction.hpp:84:48: error: ‘long double’ is not a class, struct, or union type using value_type = typename T::value_type;
```

the issue is most likely a mismatch between the g++ compiler being used and the headers supplied by the `BH` package.  There are only two solutions that I know of:

*  Upgrade g++
*  Downgrade the version of the `BH` package you are using.  The appropriate package version depends on the version of the g++ compiler you are using.

## References
<div id="refs"></div>
